---
title: "Kelp Canopy Cover Summary Using Administrative Kelp Bed Shapefiles"
author: "Gina Contolini"
date: Updated `r format(Sys.time(), '%d %B %Y')`
output: 
  html_document:
    fig_crop: FALSE
header-includes:
- \usepackage{caption}
- \captionsetup[figure]{labelformat=empty}
toc: FALSE
        
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, warning=FALSE)
options(knitr.table.format = "latex")
# to create a PDF output, change "output:html_document" to "output: pdf_document" in the header above.
# note about fig_crop:FALSE. This makes the PDF output look nicer by putting spaces btwn the figures instead of smashing them as close together as possible. It doesn't do anything to the HTML output as far as I can tell.
# how to assign username: git config --global user.name "gcontolini"
# how to add, commit, and add message in terminal: git commit -am "commit message"
# how to write commit message in terminal: write message then press esc and :wq to save and exit

#### Load libraries ####
library(ggplot2) # plotting
library(ggpubr) # arranging plots
library(scales) # let's you put commas in the axis labels for thousands
library(dplyr) # data manipulations
library(tidyr) # data cleaning
library(readr) # better data reading
library(ncdf4)
library(sf) # special features. probably mapping. 
library(sp) # spatial data
library(lubridate) # manipulating dates easily
library(knitr) # to knit the Rmd file
library(kableExtra) # pretty tables
library(bookdown) # pretty table additions

#### ggplot theme ####
my.theme = theme( 
   panel.background = element_blank(), 
   panel.grid.minor.x = element_blank(),
   panel.grid.major.x = element_blank(),
   panel.grid = element_blank(),
   axis.text  = element_text(size=10, face='bold', color='black'), #makes axis labels bigger
   axis.title = element_text(size=10, face="bold"), #makes axis title bigger and bold
   #axis.title.y = element_text(margin=margin(0,10,0,5)), #pushes the y axis title away from the y axis labels
   strip.background = element_blank(),
   #legend.title = element_blank(),
   #legend.text = element_text(size=8),
   legend.background = element_blank(), 
   legend.key = element_blank(), #makes the legend background behind the symbols go away.
   #axis.ticks.x = element_blank()
   axis.line = element_line(color = 'black'), 
   panel.border = element_blank(), # this removes the box around the plot   
   plot.title = element_text(size=12 ,face='bold', hjust = 0.5), # hjust 0.5 makes title centered
   axis.text.x = element_text(size = 10, angle = 90, vjust = 0.5, margin = margin(0,0,5,0)) # rotates x axis labels 90 degrees counter-clockwise, pushes bars onto x axis
)


#### Import data ####
# info on the data: https://sbclter.msi.ucsb.edu/data/catalog/package/?package=knb-lter-sbc.74

# Instructions from Dr. Tom Bell:
# “The output files show the Admin Bed # in the first column and the rest of the columns are quarters from 1984 - 2020. Area is in m2 and biomass is in fresh kg. Missing values (due to lack of imagery or clouds are labelled as -999).”

# Edible seaweed data from CDFW
# edible = read.csv('EdibleSeaweedHarvest_rfm_1-28-21.csv') # edible seaweed harvest. This one has months. That is important.
# kelp = read.csv('KelpHarvest_vers2_rfm_1-28-21.csv') # kelp harvest
```
For more reports like this, visit my RPubs user profile: https://rpubs.com/gcontolini

# Objective
Summarize California kelp abundance trends from Landsat satellite imagery (SBC LTER et al. 2022) by Administrative Kelp Bed (AKB) using CDFW AKB shapefiles, available at: https://map.dfg.ca.gov/arcgis/rest/services/Project_Marine/Marine_Management/MapServer/13.

# Methods
## Satellite imagery
Kelp canopy area data are from the Santa Barbara Coastal Long-term Ecological Research Landsat satellite data series updated through all four quarters of 2021 (Santa Barbara Coastal LTER et al. 2022; https://portal.edirepository.org/nis/mapbrowse?scope=knb-lter-sbc&identifier=74&revision=15). The data represents a time series of canopy area of giant kelp, *Macrocystis pyrifera*, and bull kelp, *Nereocystis luetkeana* derived from Landsat 5 Thematic Mapper (TM), Landsat 7 Enhanced Thematic Mapper Plus (ETM+), and Landsat 8 Operational Land Imager (OLI) satellite imagery, along with relevant metadata. The kelp canopy is composed of the portions of fronds and stipes floating on the surface of the water. Canopy area (m) data are given for individual 30 x 30 meter pixels for all coastal areas of California, including the Northern and Southern Channel Islands.

Data were derived from the three Landsat sensors listed above. Observations are made on a 16 day repeat cycle for each instrument, but the temporal coverage is irregular because of cloud cover, instrument failure, and the mission length of each sensor (TM: 1984 – 2011, ETM+: 1999 – present, OLI: 2013 – present). Estimates of canopy area are derived from the fractional cover of kelp canopy determined from satellite surface reflectance. Updates to the 2021 dataset include correcting cloud reflectance errors and improving the landmasking procedure, which increased the detection of kelp especially in the northern region. The different Landsat sensors were calibrated to each other using simulated Landsat data derived from hyperspectral imagery. Missing data due to the ETM+ scan line corrector error were filled using a synchrony-based gap filling method. 

Data are organized into a single NetCDF file and contain the quarterly area means for each Landsat pixel across the three sensors. Relevant metadata such as number of Landsat estimates from which the mean was derived, the number of estimates from each sensor, standard error for each quarterly estimate, spatial coordinates, and date are all included in the file.

## Plotting
Figures were created using the above dataset in conjunction with CDFW Administrative Kelp Bed shapefiles. 

```{r Load Landsast data into AKB shapefiles, include = FALSE}
# ## THIS HAS ALL BEEN COMMENTED BECAUSE IT IS JUST THE WORK THAT LED TO THE CSVS I'M USING FOR THE FIGS
# 
# # info on the data: https://sbclter.msi.ucsb.edu/data/catalog/package/?package=knb-lter-sbc.74
# 
# #Kelp Biomass and Area Dynamics of CDFW Admin Beds
# #Tom Bell
# #Novemeber 18, 2020
# # **Please let me know if you need help with any step!**
# 
# #Load Admin Bed Shapefile
# #Place the lat/lon version of the Admin Bed shapefile in your working directory
# # admin_beds <- st_read('KelpAdmin_latlong/KelpAdmin_latllong.shp')
# admin_beds = st_read('MAN_CA_KelpAdmin.shx') # wait I changed it to .shx and now it works??
# counties <- st_read('MAN_CA_CountyStateWaters.shx')
# 
# 
# #Open Kelp NetCDF File
# #Place the NetCDF file in your working directory
# #The file can be found here: https://sbclter.msi.ucsb.edu/data/catalog/package/?package=knb-lter-sbc.74
# # nc_data_2021 <- nc_open('kelpCanopyFromLandsat_2021.nc') # old data
# nc_data <- nc_open('LandsatkelpBiomass_2021_v2_withmetadata.nc') # through 2021 data
# 
# #Load necessary variables from NetCDF File
# lon <- ncvar_get(nc_data, 'longitude')
# lat <- ncvar_get(nc_data, 'latitude')
# years <- ncvar_get(nc_data, 'year')
# quarters <- ncvar_get(nc_data, 'quarter')
# biomass <- ncvar_get(nc_data, 'biomass')
# area <- ncvar_get(nc_data, 'area')
# 
# #Make Coordinate Dataframe
# coords <- data.frame(lon,lat)
# 
# #Create output matrices
# LN = dim(quarters)
# output_biomass <- matrix(ncol=LN, nrow=87)
# output_area <- matrix(ncol=LN, nrow=87)
# 
# for(i in 1:87){
# 
#   #Pull shapefile coordinates from each admin bed
#   #I had to make hard code an addition here for Bed 86 as I am not as fluent in geospatial analysis in R as I am in Matlab
#   #The issue is that the shaprefile geometry is divided into multiple parts when there are NaNs in the lat/lon columns
#   if (i == 86) {
#     bed_coords <- as.data.frame(admin_beds[i,]$geometry[[1]][[2]][[1]])
#   } else {
#     bed_coords <- as.data.frame(admin_beds[i,]$geometry[[1]][[1]][[1]])
#   }
# 
#   #Find the kelp coordinates that exist inside each admin bed polygon
#   kelp_poly <- point.in.polygon(lon, lat, bed_coords$V1, bed_coords$V2, mode.checked=FALSE)
#   TF <- which(kelp_poly %in% 1)
# 
#   #Extract biomass and area data for each polygon
#   bio_bed <- biomass[TF,]
#   area_bed <- area[TF,]
# 
#   #Sum biomass and area
#   bio_bed_sum <- colSums(bio_bed, na.rm = TRUE, dims = 1)
#   area_bed_sum <- colSums(area_bed, na.rm = TRUE, dims = 1)
# 
#   #Remove Columns w/ >25% No Data (Clouds)
#   #Find all NA values (missing data) in the kelp dataset
#   bio_bed_NA = is.na(bio_bed)
#   area_bed_NA = is.na(area_bed)
# 
#   #Find number of kelp pixels in the polygon
#   L = dim(bio_bed)
# 
#   #Calculate the percent of NAs for each season
#   bio_bed_NA_percent <- colSums(bio_bed_NA, na.rm = TRUE)/L[1]
#   area_bed_NA_percent <- colSums(area_bed_NA, na.rm = TRUE)/L[1]
# 
#   #Find seasons with greater than 20% missing data
#   NA_GT <- which(bio_bed_NA_percent > 0.25)
#   NA_GTA <- which(area_bed_NA_percent > 0.25)
# 
#   #Set seasons with greater than 20% missing data to -999
#   bio_bed_sum[NA_GT] = -999
#   area_bed_sum[NA_GTA] = -999
# 
#   #Add summed kelp canopy data to matrix
#   output_biomass[i,] <- bio_bed_sum
#   output_area[i,] <- area_bed_sum
# 
# }
# 
# #Transpose Quarter and Year data
# quarterst <- t(quarters)
# yearst <- t(years)
# 
# #Add Quarter and Year data to output matrices
# output2_biomass <- rbind(quarterst, output_biomass)
# output3_biomass <- rbind(yearst, output2_biomass)
# output2_area <- rbind(quarterst, output_area)
# output3_area <- rbind(yearst, output2_area)
# 
# #Add Admin Bed Number to output matrices
# BedsNA <- admin_beds$KelpBed
# Beds <- c(rep(NA,2),BedsNA)
# output4_biomass <- cbind(Beds, output3_biomass)
# output4_area <- cbind(Beds, output3_area)
# 
# #Write CSV of kelp biomass time series for each Admin Bed
# write.csv(output4_biomass, 'kelp_biomass_output_2022.csv', row.names = FALSE)
# write.csv(output4_area, 'kelp_area_output_2022.csv', row.names = FALSE)

# The above just organizes data into AKBs, which can only be done in a directory that can hold the large NetCDF file (>100 MB), which RStudio Cloud cannot. So I ran that code on my local machine and now I will read in the resulting .csv:

output4_area <- read.csv('kelp_area_output_2022.csv')

# I believe we only got countywide data directly from T Bell.
```

```{r admin kelp county canopy area tidying, include = FALSE}
canopy_area <- output4_area %>%
   as.data.frame()

# I want to plot area vs. year.quarter. This means year.quarter needs to be in a column. Currently all cols are separate. This is called wide format. I want tall format.

# create column names
names(canopy_area) <- paste(canopy_area[1,], canopy_area[2,], sep='.') # paste years with quarters to make each col have a unique name.

# rename the ID col for counties
names(canopy_area)[1] <- "AKB"

# remove first 2 rows since they are not canopy data, just IDs--year and quarter
canopy_area <- canopy_area[-c(1:2),] 

# convert to tall format
canopy_area_tall <- canopy_area %>% 
   pivot_longer(-'AKB', names_to = 'year.qtr', values_to = 'm2')

# separate year and quarter into two cols
canopy_area_tall <- canopy_area_tall %>%
   separate(col = year.qtr, sep = "[.]", into = c("year","qtr")) # omg. you have to hard bracket the period for some reason. wtf does [^[:alnum:]]+ mean and why aren't brackets used in the example?????? :((

# make number cols numeric
canopy_area_tall$year <- as.numeric(canopy_area_tall$year)
canopy_area_tall$qtr <- as.numeric(canopy_area_tall$qtr)
canopy_area_tall$m2 <- as.numeric(canopy_area_tall$m2)

# make quarters actual quarter decimals for plotting
qtrs <- seq(0,0.75,0.25)
for(i in 1:4) {
   canopy_area_tall$qtr[which(canopy_area_tall$qtr == i)] <- qtrs[i]
}

# add year and quarter to get a continuous time column. Note that the tibble hides the decimals place for no apparent reason but the sum is correct.
canopy_area_tall$year.qtr <- canopy_area_tall$year + canopy_area_tall$qtr

# Change negative area to NA
canopy_area_tall$m2[which(canopy_area_tall$m2 < 0)] = NA

# Convert m2 to km2
canopy_area_tall$km2 = canopy_area_tall$m2 / 1e6
```

# Figures
```{r kelp canopy area by AKB, fig.height = 3, fig.width = 10}

# All quarters
expand.a.little = 0.02

## AKB 312, Crescent City
ggplot(canopy_area_tall[which(canopy_area_tall$AKB == '312'),], aes(x = year.qtr, y = km2)) +
   geom_point() +
   geom_line() +
   scale_x_continuous(breaks = 1984:2021, expand = c(expand.a.little,0)) +
   scale_y_continuous(expand = c(expand.a.little, 0)) +
   labs(x = '',  y =  bquote('Canopy area'~(km^2))) +
   ggtitle('AKB 312 Quarterly') +
   my.theme

## AKB 311
ggplot(canopy_area_tall[which(canopy_area_tall$AKB == '311'),], aes(x = year.qtr, y = km2)) +
   geom_point() +
   geom_line() +
   scale_x_continuous(breaks = 1984:2021, expand = c(expand.a.little,0)) +
   scale_y_continuous(expand = c(expand.a.little, 0)) +
   labs(x = '',  y =  bquote('Canopy area'~(km^2))) +
   ggtitle('AKB 311 Quarterly') +
   my.theme

## AKB 310
ggplot(canopy_area_tall[which(canopy_area_tall$AKB == '310'),], aes(x = year.qtr, y = km2)) +
   geom_point() +
   geom_line() +
   scale_x_continuous(breaks = 1984:2021, expand = c(expand.a.little,0)) +
   scale_y_continuous(expand = c(expand.a.little, 0)) +
   labs(x = '',  y =  bquote('Canopy area'~(km^2))) +
   ggtitle('AKB 310 Quarterly') +
   my.theme

## AKB 309
ggplot(canopy_area_tall[which(canopy_area_tall$AKB == '309'),], aes(x = year.qtr, y = km2)) +
   geom_point() +
   geom_line() +
   scale_x_continuous(breaks = 1984:2021, expand = c(expand.a.little,0)) +
   scale_y_continuous(expand = c(expand.a.little, 0)) +
   labs(x = '',  y =  bquote('Canopy area'~(km^2))) +
   ggtitle('AKB 309 Quarterly') +
   my.theme

## AKB 308
ggplot(canopy_area_tall[which(canopy_area_tall$AKB == '308'),], aes(x = year.qtr, y = km2)) +
   geom_point() +
   geom_line() +
   scale_x_continuous(breaks = 1984:2021, expand = c(expand.a.little,0)) +
   scale_y_continuous(breaks = seq(0, 0.04, 0.01), expand = c(expand.a.little,0)) +
   labs(x = '',  y =  bquote('Canopy area'~(km^2))) +
   ggtitle('AKB 308 Quarterly') +
   my.theme

## AKB 307, Fort Bragg
ggplot(canopy_area_tall[which(canopy_area_tall$AKB == '307'),], aes(x = year.qtr, y = km2)) +
   geom_point() +
   geom_line() +
   scale_x_continuous(breaks = 1984:2021, expand = c(expand.a.little,0)) +
   scale_y_continuous(expand = c(expand.a.little, 0)) +
   labs(x = '',  y =  bquote('Canopy area'~(km^2))) +
   ggtitle('AKB 307 Quarterly') +
   my.theme

## AKB 306
ggplot(canopy_area_tall[which(canopy_area_tall$AKB == '306'),], aes(x = year.qtr, y = km2)) +
   geom_point() +
   geom_line() +
   scale_x_continuous(breaks = 1984:2021, expand = c(expand.a.little,0)) +
   scale_y_continuous(expand = c(expand.a.little, 0)) +
   labs(x = '',  y =  bquote('Canopy area'~(km^2))) +
   ggtitle('AKB 306 Quarterly') +
   my.theme

## AKB 305
ggplot(canopy_area_tall[which(canopy_area_tall$AKB == '305'),], aes(x = year.qtr, y = km2)) +
   geom_point() +
   geom_line() +
   scale_x_continuous(breaks = 1984:2021, expand = c(expand.a.little,0)) +
   scale_y_continuous(expand = c(expand.a.little, 0)) +
   labs(x = '',  y =  bquote('Canopy area'~(km^2))) +
   ggtitle('AKB 305 Quarterly') +
   my.theme

## AKB 304
ggplot(canopy_area_tall[which(canopy_area_tall$AKB == '304'),], aes(x = year.qtr, y = km2)) +
   geom_point() +
   geom_line() +
   scale_x_continuous(breaks = 1984:2021, expand = c(expand.a.little,0)) +
   scale_y_continuous(expand = c(expand.a.little, 0)) +
   labs(x = '',  y =  bquote('Canopy area'~(km^2))) +
   ggtitle('AKB 304 Quarterly') +
   my.theme

## AKB 303
ggplot(canopy_area_tall[which(canopy_area_tall$AKB == '303'),], aes(x = year.qtr, y = km2)) +
   geom_point() +
   geom_line() +
   scale_x_continuous(breaks = 1984:2021, expand = c(expand.a.little,0)) +
   scale_y_continuous(expand = c(expand.a.little, 0)) +
   labs(x = '',  y =  bquote('Canopy area'~(km^2))) +
   ggtitle('AKB 303 Quarterly') +
   my.theme

## AKB 302
ggplot(canopy_area_tall[which(canopy_area_tall$AKB == '302'),], aes(x = year.qtr, y = km2)) +
   geom_point() +
   geom_line() +
   scale_x_continuous(breaks = 1984:2021, expand = c(expand.a.little,0)) +
   scale_y_continuous(expand = c(expand.a.little, 0)) +
   labs(x = '',  y =  bquote('Canopy area'~(km^2))) +
   ggtitle('AKB 302 Quarterly') +
   my.theme

## AKB 301
ggplot(canopy_area_tall[which(canopy_area_tall$AKB == '301'),], aes(x = year.qtr, y = km2)) +
   geom_point() +
   geom_line() +
   scale_x_continuous(breaks = 1984:2021, expand = c(expand.a.little,0)) +
   scale_y_continuous(expand = c(expand.a.little, 0)) +
   labs(x = '',  y =  bquote('Canopy area'~(km^2))) +
   ggtitle('AKB 301 Quarterly') +
   my.theme
```

Figure 1. Quarterly canopy area in Administrative Kelp Beds as estimated from Landsat satellites (Santa Barbara Coastal LTER et al. 2022). Canopy area does not distinguish between bull kelp and giant kelp.  Missing values indicate no data is available. Data is updated through the last quarter of 2021.

# Tables

```{r exploring which quarters have the annual maximum, echo = FALSE}
qtr_max_AKB <- canopy_area_tall %>%
  group_by(AKB, year) %>%        # group by AKB and year
  slice(which.max(km2)) %>%      # find the maximum km2 value
  subset(select = -year.qtr) %>% # remove year.qtr col
  transmute(quarter = qtr*4+1)   # change quarters from decimals to 1,2,3,4

# Count how many quarters are the max per year per AKB
n_max_qtrs_AKB <- qtr_max_AKB %>%
  group_by(AKB, quarter) %>%
  summarise(number = length(quarter), .groups = 'drop') %>%
  filter(AKB > 300) %>%
  complete(AKB, quarter, fill = list(number = 0)) # sets missing quarters to number = 0

# kable(n_max_qtrs_AKB
#       , digits = 2
#       , booktabs = TRUE
#       , row.names = FALSE
#       , align = 'c'
#       , linesep = ''
#       , caption = 'Number of quarters with maximum annual canopy area'
# ) %>%
#   kable_styling(latex_options = 'hold_position') # otherwise the table moves in the pdf

# which quarters had the greatest canopy area in 2021?
# I want to identify the quarter in each of the 300-level AKBs that had the max km2 value in 2021
max_2021_km2 <- canopy_area_tall %>%
  filter(year == 2021) %>%
  filter(AKB > 300) %>% # this shows km2 values for all 300-level AKBs in 2021. What is the max for 2021?
  group_by(AKB) %>%
  summarise(max.km2 = max(km2, na.rm = T)) # this gives the max km2 values. which quarter are they in?
      
km2_qtrs_300AKBs <- canopy_area_tall %>%
  filter(year == 2021 & AKB > 300) %>%
  #filter(km2 %in% max_2021_km2$max.km2) %>% # This shows only the quarters with the annual maxima
  mutate(quarter = qtr*4+1) %>%
  subset(select = c('AKB','year','quarter','km2')) %>%
  arrange(desc(AKB)) %>%
  relocate(year) # puts year as the first col

kable(km2_qtrs_300AKBs
      #, digits = 2
      , booktabs = TRUE
      , row.names = FALSE
      , align = 'c'
      , linesep = ''
      , caption = 'Quarterly kelp canopy area in 2021 only'
) %>%
  kable_styling(latex_options = 'hold_position') # otherwise the table moves in the pdf

```


# Literature Cited
Santa Barbara Coastal LTER, T. Bell, K. Cavanaugh, and D. Siegel. 2022. SBC LTER: Time series of quarterly NetCDF files of kelp biomass in the canopy from Landsat 5, 7 and 8, since 1984 (ongoing) ver 14. Environmental Data Initiative. https://doi.org/10.6073/pasta/89b63c4b49b80fb839613e9d389d9902 (Accessed 2021-01-26).