---
title: "Kelp Canopy Cover Summary Using CDFW County Shapefiles"
author: "Gina Contolini"
date: Updated `r format(Sys.time(), '%d %B %Y')`
output: 
  html_document:
    fig_crop: FALSE
header-includes:
- \usepackage{caption}
- \captionsetup[figure]{labelformat=empty}
toc: FALSE
        
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, warning=FALSE)
options(knitr.table.format = "latex")
# to create a PDF output, change "output:html_document" to "output: pdf_document" in the header above.
# note about fig_crop:FALSE. This makes the PDF output look nicer by putting spaces btwn the figures instead of smashing them as close together as possible. It doesn't do anything to the HTML output as far as I can tell.
# how to assign username: git config --global user.name "gcontolini"
# how to add, commit, and add message in terminal: git commit -am "commit message"
# how to write commit message in terminal: write message then press esc and :wq to save and exit

#### Load libraries ####
library(ggplot2) # plotting
library(ggpubr) # arranging plots
library(scales) # let's you put commas in the axis labels for thousands
library(dplyr) # data manipulations
library(tidyr) # data cleaning
library(readr) # better data reading
library(ncdf4)
library(sf) # special features. probably mapping. 
library(sp) # spatial data
library(lubridate) # manipulating dates easily
library(knitr) # to knit the Rmd file
library(kableExtra) # pretty tables
library(bookdown) # pretty table additions

#### ggplot theme ####
my.theme = theme( 
   panel.background = element_blank(), 
   panel.grid.minor.x = element_blank(),
   panel.grid.major.x = element_blank(),
   panel.grid = element_blank(),
   axis.text  = element_text(size=10, face='bold', color='black'), #makes axis labels bigger
   axis.title = element_text(size=10, face="bold"), #makes axis title bigger and bold
   #axis.title.y = element_text(margin=margin(0,10,0,5)), #pushes the y axis title away from the y axis labels
   strip.background = element_blank(),
   #legend.title = element_blank(),
   #legend.text = element_text(size=8),
   legend.background = element_blank(), 
   legend.key = element_blank(), #makes the legend background behind the symbols go away.
   #axis.ticks.x = element_blank()
   axis.line = element_line(color = 'black'), 
   panel.border = element_blank(), # this removes the box around the plot   
   plot.title = element_text(size=12 ,face='bold', hjust = 0.5), # hjust 0.5 makes title centered
   axis.text.x = element_text(size = 10, angle = 90, vjust = 0.5, margin = margin(0,0,5,0)) # rotates x axis labels 90 degrees counter-clockwise, pushes bars onto x axis
)


#### Import data ####
# info on the data: https://sbclter.msi.ucsb.edu/data/catalog/package/?package=knb-lter-sbc.74

# Instructions from Dr. Tom Bell:
# “The output files show the Admin Bed # in the first column and the rest of the columns are quarters from 1984 - 2020. Area is in m2 and biomass is in fresh kg. Missing values (due to lack of imagery or clouds are labelled as -999).”

# Edible seaweed data from CDFW
# edible = read.csv('EdibleSeaweedHarvest_rfm_1-28-21.csv') # edible seaweed harvest. This one has months. That is important.
# kelp = read.csv('KelpHarvest_vers2_rfm_1-28-21.csv') # kelp harvest
```

# Objective
Summarize California kelp abundance trends from Landsat satellite imagery (Santa Barbara Coastal LTER et al. 2022) by county using shapefiles of state waters delineated by coastal counties.

# Methods
## Satellite imagery
Kelp canopy area data are from the Santa Barbara Coastal Long-term Ecological Research Landsat satellite data series updated through all four quarters of 2021 (Santa Barbara Coastal LTER et al. 2022; https://doi.org/10.6073/pasta/89b63c4b49b80fb839613e9d389d9902). The data represents a time series of canopy area of giant kelp, *Macrocystis pyrifera*, and bull kelp, *Nereocystis luetkeana* derived from Landsat 5 Thematic Mapper (TM), Landsat 7 Enhanced Thematic Mapper Plus (ETM+), and Landsat 8 Operational Land Imager (OLI) satellite imagery, along with relevant metadata. The kelp canopy is composed of the portions of fronds and stipes floating on the surface of the water. Canopy area (m) data are given for individual 30 x 30 meter pixels for all coastal areas of California, including the Northern and Southern Channel Islands.

Data were derived from the three Landsat sensors listed above. Observations are made on a 16 day repeat cycle for each instrument, but the temporal coverage is irregular because of cloud cover, instrument failure, and the mission length of each sensor (TM: 1984 – 2011, ETM+: 1999 – present, OLI: 2013 – present). Estimates of canopy area are derived from the fractional cover of kelp canopy determined from satellite surface reflectance. Updates to the 2021 dataset include correcting cloud reflectance errors and improving the landmasking procedure, which increased the detection of kelp especially in the northern region. The different Landsat sensors were calibrated to each other using simulated Landsat data derived from hyperspectral imagery. Missing data due to the ETM+ scan line corrector error were filled using a synchrony-based gap filling method. 

Data are organized into a single NetCDF file and contain the quarterly area means for each Landsat pixel across the three sensors. Relevant metadata such as number of Landsat estimates from which the mean was derived, the number of estimates from each sensor, standard error for each quarterly estimate, spatial coordinates, and date are all included in the file.

## Plotting
Figures were created using the above dataset in conjunction with CDFW county shapefiles, which include state waters delineated by coastal counties. Maximum annual canopy area is defined as the quarter each year with the greatest canopy area.

```{r Load county shapefiles and match with Landsast data, include = FALSE}
# The loop below requires the NetCDF file which is too large for RStudio Cloud to host. Also takes 5 minutes to run. I commented it out, ran it on my local machine, and read in the output at the end.

# #Kelp Biomass and Area Dynamics of State Waters Delineated by Coastal Counties
# #Tom Bell
# #November 18, 2020
# # **Please let me know if you need help with any step!**
# 
# #Install Packages **(Only the first time you ever run this script)**
# #The first time you ever run this script, please install these packages by removing the '#' in front of each line (lines 8 - 10)
# # install.packages('ncdf4')
# # install.packages('sf')
# # install.packages('sp')
# 
# #Load Required Packages
# #You need to load these packages each time you restart an R session and run this script
# library(ncdf4)
# library(sf)
# library(sp)
# 
# #Set Work Directory
# #This is a folder on your computer where all input files and outputs files will be placed
# # **Please update this path to the working directory on your computer**
# # **Note that on a Windows machine the forward slashes "/" have to be changed to back slashes "\"
# setwd("C:/Users/gcontolini/OneDrive - California Department of Fish and Wildlife/Documents/Bull Kelp Working Group/Data analysis")
# 
# #Load Coastal Counties State Waters Shapefile
# #Place the lat/lon version of the Coastal Counties State Waters shapefile in your working directory
# county_shapes <- st_read('MAN_CA_CountyStateWaters.shp')
# 
# #Open Kelp NetCDF File
# #Place the NetCDF file in your working directory
# #The file can be found here: https://sbclter.msi.ucsb.edu/data/catalog/package/?package=knb-lter-sbc.74
# nc_data <- nc_open('LandsatkelpBiomass_2021_v2_withmetadata.nc')
# 
# #Load necessary variables from NetCDF File
# lon <- ncvar_get(nc_data, 'longitude')
# lat <- ncvar_get(nc_data, 'latitude')
# years <- ncvar_get(nc_data, 'year')
# quarters <- ncvar_get(nc_data, 'quarter')
# biomass <- ncvar_get(nc_data, 'biomass')
# area <- ncvar_get(nc_data, 'area')
# 
# #Make Coordinate Dataframe
# coords <- data.frame(lon,lat)
# 
# #Create output matrices
# LN = dim(quarters)
# output_biomass <- matrix(ncol=LN, nrow=22)
# output_area <- matrix(ncol=LN, nrow=22)
# 
# for(i in 1:22){
#   
#   #Pull shapefile coordinates from each coastal county
#   if (i == 17) {
#     county_coords <- as.data.frame(county_shapes[i,]$geometry[[1]][[30]][[1]])
#   } else {
#     county_coords <- as.data.frame(county_shapes[i,]$geometry[[1]][[1]][[1]])
#   }
# 
#   #Find the kelp coordinates that exist inside each county polygon
#   kelp_poly <- point.in.polygon(lon, lat, county_coords$V1, county_coords$V2, mode.checked=FALSE)
#   TF <- which(kelp_poly %in% 1)
# 
#   #Extract biomass and area data for each polygon
#   bio_county <- biomass[TF,] 
#   area_county <- area[TF,] 
# 
#   #Sum biomass and area
#   bio_county_sum <- colSums(bio_county, na.rm = TRUE, dims = 1)
#   area_county_sum <- colSums(area_county, na.rm = TRUE, dims = 1)
#   
#   #Remove Columns w/ >25% No Data (Clouds)
#   #Find all NA values (missing data) in the kelp dataset
#   bio_county_NA = is.na(bio_county)
#   area_county_NA = is.na(area_county)
#   
#   #Find number of kelp pixels in the polygon
#   L = dim(bio_county)
#   
#   #Calculate the percent of NAs for each season
#   bio_county_NA_percent <- colSums(bio_county_NA, na.rm = TRUE)/L[1]
#   area_county_NA_percent <- colSums(area_county_NA, na.rm = TRUE)/L[1]
#   
#   #Find seasons with greater than 20% missing data
#   NA_GT <- which(bio_county_NA_percent > 0.25)
#   NA_GTA <- which(area_county_NA_percent > 0.25)
#   
#   #Set seasons with greater than 20% missing data to -999
#   bio_county_sum[NA_GT] = -999
#   area_county_sum[NA_GTA] = -999
# 
#   #Add summed kelp canopy data to matrix
#   output_biomass[i,] <- bio_county_sum
#   output_area[i,] <- area_county_sum
#   
# }
# 
# #Transpose Quarter and Year data
# quarterst <- t(quarters)
# yearst <- t(years)
# 
# #Add Quarter and Year data to output matrices
# output2_biomass <- rbind(quarterst, output_biomass)
# output3_biomass <- rbind(yearst, output2_biomass)
# output2_area <- rbind(quarterst, output_area)
# output3_area <- rbind(yearst, output2_area)
# 
# #Add County name to output matrices
# CountyNA <- county_shapes$County
# Counties <- c(rep(NA,2),CountyNA)
# output4_biomass <- cbind(Counties, output3_biomass)
# output4_area <- cbind(Counties, output3_area)
# 
# #Write CSV of kelp biomass time series for each county
# write.csv(output4_biomass, 'kelp_biomass_county_output.csv', row.names = FALSE)
# write.csv(output4_area, 'kelp_area_county_output.csv', row.names = FALSE)


output4_area <- read_csv("kelp_area_county_output_01Feb2022.csv")

```

```{r admin kelp county canopy area tidying, include = FALSE}
canopy_area <- output4_area %>%
   as.data.frame()

# I want to plot area vs. year.quarter. This means year.quarter needs to be in a column. Currently all cols are separate. This is called wide format. I want tall format.

# create column names
names(canopy_area) <- paste(canopy_area[1,], canopy_area[2,], sep='.') # paste years with quarters to make each col have a unique name.

# rename the ID col for counties
names(canopy_area)[1] <- "county"

# remove first 2 rows since they are not canopy data, just IDs--year and quarter
canopy_area <- canopy_area[-c(1:2),] 

# convert to tall format
canopy_area_tall <- canopy_area %>% 
   pivot_longer(-'county', names_to = 'year.qtr', values_to = 'm2')

# separate year and quarter into two cols
canopy_area_tall <- canopy_area_tall %>%
   separate(col = year.qtr, sep = "[.]", into = c("year","qtr")) # omg. you have to hard bracket the period for some reason. wtf does [^[:alnum:]]+ mean and why aren't brackets used in the example?????? :((

# make number cols numeric
canopy_area_tall$year <- as.numeric(canopy_area_tall$year)
canopy_area_tall$qtr <- as.numeric(canopy_area_tall$qtr)
canopy_area_tall$m2 <- as.numeric(canopy_area_tall$m2)

# make quarters actual quarter decimals for plotting
qtrs <- seq(0,0.75,0.25)
for(i in 1:4) {
   canopy_area_tall$qtr[which(canopy_area_tall$qtr == i)] <- qtrs[i]
}

# add year and quarter to get a continuous time column. Note that the tibble hides the decimals place for no apparent reason but the sum is correct.
canopy_area_tall$year.qtr <- canopy_area_tall$year + canopy_area_tall$qtr

# Change negative area to NA
canopy_area_tall$m2[which(canopy_area_tall$m2 < 0)] = NA

# Convert m2 to km2
canopy_area_tall$km2 = canopy_area_tall$m2 / 1e6

```


# Figures
```{r quarterly canopy area in each county, fig.height = 3}
# All quarters
expand.a.little = 0.02

##Del Norte
ggplot(canopy_area_tall[which(canopy_area_tall$county == 'Del Norte'),], aes(x = year.qtr, y = km2)) +
   geom_point() +
   geom_line() +
   scale_x_continuous(breaks = 1984:2021, expand = c(expand.a.little,0)) +
   # scale_y_continuous(breaks = seq(0, 0.015, 0.0025), expand = c(expand.a.little,0)) +
   labs(x = '',  y =  bquote('Canopy area'~(km^2))) +
   ggtitle('Del Norte County Quarterly') +
   my.theme

## Humboldt
ggplot(canopy_area_tall[which(canopy_area_tall$county == 'Humboldt'),], aes(x = year.qtr, y = km2)) +
   geom_point() +
   geom_line() +
   scale_x_continuous(breaks = 1984:2021, expand = c(expand.a.little,0)) +
   # scale_y_continuous(limits = c(0,0.0225), breaks = seq(0, 0.025, 0.0025), expand = c(expand.a.little,0)) +
   labs(x = '',  y =  bquote('Canopy area'~(km^2))) +
   ggtitle('Humboldt County Quarterly') +
   my.theme

## Mendocino
ggplot(canopy_area_tall[which(canopy_area_tall$county == 'Mendocino'),], aes(x = year.qtr, y = km2)) +
   geom_point() +
   geom_line() +
   scale_x_continuous(breaks = 1984:2021, expand = c(expand.a.little,0)) +
   # scale_y_continuous(limits = c(0,3), breaks = seq(0, 3, 0.25), expand = c(expand.a.little,0)) +
   labs(x = '', y =  bquote('Canopy area'~(km^2))) +
   ggtitle('Mendocino County Quarterly') +
   my.theme

## Sonoma
ggplot(canopy_area_tall[which(canopy_area_tall$county == 'Sonoma'),], aes(x = year.qtr, y = km2)) +
   geom_point() +
   geom_line() +
   scale_x_continuous(breaks = 1984:2021, expand = c(expand.a.little,0)) +
   # scale_y_continuous(limits = c(0,2), breaks = seq(0, 2, 0.25), expand = c(expand.a.little,0)) +
   labs(x = '', y =  bquote('Canopy area'~(km^2))) +
   ggtitle('Sonoma County Quarterly') +
   my.theme

## Marin
ggplot(canopy_area_tall[which(canopy_area_tall$county == 'Marin'),], aes(x = year.qtr, y = km2)) +
   geom_point() +
   geom_line() +
   scale_x_continuous(breaks = 1984:2021, expand = c(expand.a.little,0)) +
   # scale_y_continuous(limits = c(0, 0.0125), breaks = seq(0, 0.0125, 0.0025), expand = c(expand.a.little,0)) +
   labs(x = '', y =  bquote('Canopy area'~(km^2))) +
   ggtitle('Marin County Quarterly') +
   my.theme
```

Figure 1. Quarterly canopy area in state waters in coastal counties as estimated from Landsat satellites (Santa Barbara Coastal LTER et al. 2022). Canopy area does not distinguish between bull kelp and giant kelp.  Missing values indicate no data is available. Data is updated through the last quarter of 2021.

```{r Annual max canopy area in each county, fig.height = 2}
# Maximum canopy area. in each year and county, find the maximum km2 value.
canopy_area_max_annual <- canopy_area_tall %>%
  group_by(county, year) %>%
  summarise(max.km2 = max(km2, na.rm = T), .groups = 'drop')

# 1987 and 1992 Marin had no data. Change -Inf to NA for those years so it plots better.
canopy_area_max_annual$max.km2[which(canopy_area_max_annual$max.km2 < 0)] <- NA

##Del Norte
ggplot(canopy_area_max_annual[which(canopy_area_max_annual$county == 'Del Norte'),], aes(x = year, y = max.km2)) +
   geom_point() +
   geom_line() +
   scale_x_continuous(breaks = 1984:2021, expand = c(expand.a.little,0)) +
   # scale_y_continuous(breaks = seq(0, 0.015, 0.0025), expand = c(expand.a.little,0)) +
   labs(x = '',  y =  bquote('Canopy area'~(km^2))) +
   ggtitle('Del Norte County Max Annual') +
   my.theme 

## Humboldt
ggplot(canopy_area_max_annual[which(canopy_area_max_annual$county == 'Humboldt'),], aes(x = year, y = max.km2)) +
   geom_point() +
   geom_line() +
   scale_x_continuous(breaks = 1984:2021, expand = c(expand.a.little,0)) +
   # scale_y_continuous(expand = c(expand.a.little,0)) +
   labs(x = '',  y =  bquote('Canopy area'~(km^2))) +
   ggtitle('Humboldt County Max Annual') +
   my.theme 

## Mendocino
ggplot(canopy_area_max_annual[which(canopy_area_max_annual$county == 'Mendocino'),], aes(x = year, y = max.km2)) +
   geom_point() +
   geom_line() +
   scale_x_continuous(breaks = 1984:2021, expand = c(expand.a.little,0)) +
   # scale_y_continuous(expand = c(expand.a.little,0)) +
   labs(x = '',  y =  bquote('Canopy area'~(km^2))) +
   ggtitle('Mendocino County Max Annual') +
   my.theme 

## Sonoma
ggplot(canopy_area_max_annual[which(canopy_area_max_annual$county == 'Sonoma'),], aes(x = year, y = max.km2)) +
   geom_point() +
   geom_line() +
   scale_x_continuous(breaks = 1984:2021, expand = c(expand.a.little,0)) +
   # scale_y_continuous(expand = c(expand.a.little,0)) +
   labs(x = '',  y =  bquote('Canopy area'~(km^2))) +
   ggtitle('Sonoma County Max Annual') +
   my.theme 

## Marin
ggplot(canopy_area_max_annual[which(canopy_area_max_annual$county == 'Marin'),], aes(x = year, y = max.km2)) +
   geom_point() +
   geom_line() +
   scale_x_continuous(breaks = 1984:2021, expand = c(expand.a.little,0)) +
   # scale_y_continuous(expand = c(expand.a.little, 0)) +
   labs(x = '',  y =  bquote('Canopy area'~(km^2))) +
   ggtitle('Marin County Max Annual') +
   my.theme 
```

Figure 2. Annual maximum canopy area (measured quarterly) in state waters in coastal counties as estimated from Landsat satellites (Santa Barbara Coastal LTER et al. 2022). Canopy area does not distinguish between bull kelp and giant kelp. Missing values indicate no data is available. Data is updated through the last quarter of 2021.

```{r Annual max canopy area in all northern counties, fig.height = 9}

``` 

# Tables

```{r which quarter has the annual maximum in each county?, echo = FALSE}
qtr_max_county <- canopy_area_tall %>%
  group_by(county, year) %>%     # group by county and year
  slice(which.max(km2)) %>%      # find the maximum km2 value
  subset(select = -year.qtr) %>% # remove year.qtr col
  transmute(quarter = qtr*4+1)   # change quarters from decimals to 1,2,3,4

# # Count how many quarters are the max per year per county
# n_max_qtrs_county <- qtr_max_county %>%
#   group_by(county, quarter) %>%
#   summarise(number = length(quarter), .groups = 'drop') %>%
#   filter(county %in% c('Del Norte', 'Humboldt', 'Mendocino', 'Sonoma')) %>%
#   complete(county, quarter, fill = list(number = 0)) # sets missing quarters to number = 0
# 
# kable(n_max_qtrs_county
#       , digits = 2
#       , booktabs = TRUE
#       , row.names = FALSE
#       , align = 'c'
#       , linesep = ''
#       , caption = 'Number of quarters with maximum annual canopy area'
# ) %>%
#   kable_styling(latex_options = 'hold_position') # otherwise the table moves in the pdf

# I want to identify the quarter in each of the counties that had the max km2 value in 2021
max_2021_km2 <- canopy_area_tall %>%
  filter(year == 2021) %>%
  filter(county %in% c('Del Norte','Humboldt','Mendocino','Sonoma','Marin')) %>% # this shows km2 values for only northern counties in 2021. What is the max for 2021?
  group_by(county) %>%
  summarise(max.km2 = max(km2, na.rm = T)) # this gives the max km2 values. which quarter are they in?
      
km2_qtrs_northcounties <- canopy_area_tall %>%
  filter(year == 2021 & county %in% c('Del Norte','Humboldt','Mendocino','Sonoma','Marin')) %>%
  #filter(km2 %in% max_2021_km2$max.km2) %>% # This shows only the quarters with the annual maxima
  mutate(quarter = qtr*4+1) %>%
  subset(select = c('county','year','quarter','km2')) %>%
  arrange(county) %>% # order by county name north to south
  relocate(year) # puts year as the first col


kable(km2_qtrs_northcounties
      #, digits = 2
      , booktabs = TRUE
      , row.names = FALSE
      , align = 'c'
      , linesep = ''
      , caption = 'Quarterly kelp canopy area in 2021 only'
) %>%
  kable_styling(latex_options = 'hold_position') # otherwise the table moves in the pdf


      
```


# Literature Cited
Santa Barbara Coastal LTER, T. Bell, K. Cavanaugh, and D. Siegel. 2022. SBC LTER: Time series of quarterly NetCDF files of kelp biomass in the canopy from Landsat 5, 7 and 8, since 1984 (ongoing) ver 14. Environmental Data Initiative. Updated 16 Jan 2022. https://doi.org/10.6073/pasta/89b63c4b49b80fb839613e9d389d9902 (Accessed 2022-01-26).